Structural replication of nanoporous media using procedural noise

### 2017-02-04 ms.tex
Hyttetur work paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract} 
In this paper, we aim to develop a method for producing statistically correct simulations of nanoporous media without the use of conventional Molecular Dynamics Simulator (MDS). In short, the proposed model is a binary function that returns either true if the input position is in a void / pore, or false if within a wall. By using procedural noise, we are not restricted to generating media within a limited box, as the noise functions is continuous on all scales. It is therefore possible to validated pores at theoretically arbitrary range. In addition, this method is extremely fast, requiring very little computational power to calculate whether a particle is within a pore or not. After presenting such a model, we choose a method to measure the structural properties of nanoporous media that are created either through regular methods (e.g. LAMMPS) or with our method. The measure we selected is the radial pore size distribution $g(r)$, which is regarded as the de-facto standard method for capturing the structural properties of porous media. Using this measure, we define a full likelihood framework for estimating model parameters given any data set, and continue by validating our own framework by analysing procedural noise data sets with known input parameters. Finally, we perform a full likelihood analysis of silica $SiO^2$ data set simulated in LAMMPS, and show that our model correctly reproduces several of the internal properties of the data set, including $g(r)$, porosity and surface area.

@@

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Duis consequat turpis vel mollis accumsan. Etiam eleifend magna vel quam porta, sed porta nibh placerat. Sed in faucibus risus. Duis suscipit tincidunt leo, nec laoreet leo elementum elementum. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas. Sed eros risus, convallis sit amet turpis a, dapibus accumsan quam. Donec fringilla rhoncus turpis, quis ultrices odio vulputate elementum. Cras nec viverra urna. Phasellus laoreet tortor sed est sagittis laoreet. Nullam a volutpat ex.

Mauris vestibulum sit amet libero eget facilisis. Vivamus a nunc sed felis ultrices convallis luctus et mauris. Sed tincidunt mauris nunc, vehicula placerat eros faucibus quis. Morbi egestas, nisi quis luctus dignissim, mauris odio pulvinar tellus, non efficitur risus erat vel nunc. Cras vel felis magna. Mauris rutrum magna a interdum sodales. Nam tempor, diam non sodales tristique, nulla ipsum eleifend dui, non aliquam felis tellus id enim. Donec eget nulla leo.

Ut hendrerit cursus libero, in scelerisque tortor sollicitudin accumsan. Nam eleifend velit metus, quis volutpat justo faucibus ut. Cras ut sem in nunc fringilla egestas vitae sit amet justo. Ut vel condimentum tortor. Morbi in massa in ipsum ultrices tincidunt. Duis feugiat dignissim nisi ac ultricies. Quisque eget risus fermentum, condimentum massa ullamcorper, posuere neque. Nam congue consequat mi, vel commodo mi tempus eget.

\section{Method}
\todo{her står det nå det samme som i abstract}
In this paper, we aim to develop a method for producing statistically correct simulations of nanoporous media without the use of conventional Molecular Dynamics Simulator (MDS). In short, the proposed model is a binary function that returns either true if the input position is in a void / pore, or false if within a wall. By using procedural noise, we are not restricted to generating media within a limited box, as the noise functions is continuous on all scales. It is therefore possible to validated pores at theoretically infinite range. In addition, this method is extremely fast, requiring very little computational power to calculate whether a particle is within a pore or not.

After presenting such a model, we choose a method to measure the structural properties of nanoporous media that are created either through regular methods (e.g. LAMMPS \cite{plimpton1995fast}) or with our method. The measure we selected is the radial pore size distribution $g(r)$, which is regarded as the de-facto standard method for capturing the structural properties of porous media.

Using this measure, we define a full likelihood framework for estimating model parameters given any data set, and continue by validating our own framework by analysing procedural noise data sets with known input parameters. Finally, we perform a full likelihood analysis of

\subsection{Procedural noise}
\label{sec:perlin}

Procedural algorithmic methods for creating realistic-looking
fractal-like patterns have been around for more than 30 years. The most famous of these
algorithms is Perlin noise, first developed by Ken Perlin in 1982 for
use in the movie ``Tron''.  Perlin noise is a type of pseudo-random
gradient noise that mimics a Gaussian field, but at low computational cost. 
An improved version of Perlin noise is called Simplex
noise, which has several advantages over standard Perlin noise in
terms of scalability and speed.

\subsubsection{Perlin noise - an overview}
An overview of the current state of procedural noise functions can be
found in \cite{lagae:2010}.  Perlin noise \citep{perlin:2002} 
is a band limited repeatable pseudo-random function
from $\mathbb R^n \to \mathbb R$ that is computationally fast and has
properties which make it suitable for creating patterns that mimic
nature. Perlin noise is an approximation to Gaussian filtered noise
implemented as a pseudo-random spline. The algorithm uses a predefined
grid of gradients pointing in a random direction, where for any point
within a grid, we interpolate the gradients from the four corners. The
algorithm is as follows:
\begin{itemize}
  \item[1.] For a point, obtain the four closest gradients in the grid.
   \item [2.] For each of the gradients, calculate the dot product
     between the gradient and a vector defined by the  the distance
     between the the point and the current grid corner. 
   \item [3.] Instead of linear interpolation $F(t) = t$, use a fade function to produce smoother interpolated values. Perlin noise typically uses the following fade function: $F(t) = t ^3 (t  (6 - 15) + 10)$.
    \item [4.] Obtain the final value by linearly interpolating between the x-axis and
      use this result to linearly interpolate the y-axis. 
\end{itemize} 
Simplex noise works in a similar manner, but requires fewer
computational steps, scales better in higher ($\ge 3$) dimensions and has fewer 
anisotropic properties.

\subsubsection{Properties of procedural noise}
\begin{figure}
\includegraphics[width=.5\textwidth]{noise.eps}
\caption{Perlin noise (left) versus simplex noise (right). Note the
  straight x-y pattern in the single upper Perlin mode and the less
  anisotropic simplex mode. Lower part: combining octaves tend
  to remove these patterns.}
\label{fig:noise}
\end{figure}

\begin{figure}
\includegraphics[width=.5\textwidth]{noise_comparison.png}
%\mbox{\epsfig{file=perlinstats.eps, width=\linewidth}}
\caption{Normalized distributions of Perlin noise and simplex noise compared with a
  Gaussian distribution. Note how both noise functions produce
  near-Gaussian, but slightly different distributions.}
\label{fig:properties}
\end{figure}
Standard Perlin noise is known for having several anisotropic features
\citep{lagae:2010}. Perlin noise also gives rise to a ``textile''
pattern in the $x$ and $y$-directions, which is evident from the upper
left part of figure \ref{fig:noise}, depicting 2D Perlin noise in
a single octave. The upper-right part shows the same node for Simplex noise,
with less obvious patterns. However, when combining several octaves,
as depicted in the two lower parts, these anisotropic patterns tend to cancel
out. In any case, for observations of weak gravitational lensing, these anisotropies would occur on much smaller 
scales than the typical resolution. 
In addition, we have calculated the statistical distributions for Perlin and simplex noise
with $\sigma = 0.3$, and have plotted the results in figure \ref{fig:properties}.
Each procedural noise method has distributions that are close to normal, 
but with some asymmetrical properties.

\subsection{Multiple octaves: creating patterns}
\label{sec:octaves}
The different octaves from procedural noise functions can be combined
to produce a wide range of nature-like patterns. A generic
N-dimensional procedural function $\Phi(\vec x): \mathbb R^N \to
\mathbb R$ could be expressed as
\begin{equation}
\label{eq:procedural}
 \Phi(\vec x) = \Theta \Big(\sum_k \kappa (k) P(  f_s (\vec x,k) )) \Big),
 \end{equation}
where $f_s(\vec x,k)$ describes the frequency scale modifier (e.g. $f_s(\vec x,k) =
k\vec x$), $\kappa(k)$ is the octave amplitude (eg $\kappa(k) = 1/k$)
and $\Theta(x)$ an overall modifier (e.g. $\Theta(x) = x$ or
$\Theta(x) = \frac{1}{x}$). In a sense, $\kappa(k)$ is similar to Fourier coefficients. 
In its simplest case, summing directly
over the octaves with amplitude $1/k$ results in 
\begin{equation}
\label{eq:perlinlinear}
 \Phi(\vec x) = \sum_k \frac{1}{k} P( 2k\vec x), 
\end{equation}
and produces an 2D procedural noise pattern depicted in the lower
part of figure \ref{fig:noise}. Finally, since simplex noise both scales better with higher dimensions and produces less asymmetric effects, we choose to use this noise method over regular Perlin noise in our method.

\subsection{Model}
In this paper, we are interested in developing a model using Simplex/Perlin noise to reproduce nanoporous structures that are as statistically close to actual simulations of porous media. For a given data set and a model, we wish to estimate the parameters that best represent the data. This is performed through a full likelihood analysis, where we in addition validate the model by comparing the radial pore size distribution $g(r)$, porosity and surface area of the system.

\begin{figure*}
\includegraphics[width=.95\textwidth]{model_examples.png}
\caption{Examples of 3D simplex noise structures in nanoporous media. Left image: low persistence (large scale dominates), middle image: medium persistence (small scales starts dominating) and right image: high threshold (more atoms are discarded). }
\label{fig:model_example}
\end{figure*}
We construct a generic sum over procedural noise octaves with the expression: 
\begin{equation}
  M_1(\vec x) = \sum_{\omega=0}^{\Omega} \psi^{\omega -1}   N\big((\vec x + \sigma \cdot \omega)\nu \cdot 2^{\omega-1} \big),
\label{eq:noisemodel1}
\end{equation}
where $\Omega$ is the total amount of octaves, $\psi$ the persistence (scale falloff), $N$ a generic $N$-dimensional noise function, $\sigma$ a predefined random vector that removes anisotropic artefacts and $\nu$ the frequency modifier representing the initial largest scales in the model. For each mode, the frequency is thus multiplied, while the overall amplitude is damped by a given persistence. In the end, we simulate a large set of bulk particles without any walls, and use the noise function to remove particles that are not in a procedural void given by a threshold $\tau$. Three examples of this model are depicted in figure \ref{fig:model_example}, where we have used $\Omega=6$ octaves and the scale is $\nu=0.01$ and cut-off $\tau=0$. The left image depicts low persistence ($\psi = 0.0$), where only the first octave dominates (corresponding to large scales). The middle picture shows the same parameters with a higher persistence $(\psi = 0.5)$, where smaller scales become more dominating. In the rightmost picture, the effect of the cut-off threshold $\tau = -0.3$ is shown.

@@

Ideally, the model measure should pick up all geometrical properties that are relevant for the physical study of interest. For instance, pore size distribution and porosity are import for studying transport in porous media. Another such measure is the Distance-To-Atom (DTA).
The method used for calculating the porosity of a media is defined as in \cite{gelb1998characterization}, and we obtain the surface area of a porous media by using marching cubes triangulation to sum the surface area of the triangles.

In this paper, we choose to use the radial pore size distribution $g(r)$ as defined and implemented in LAMMPS \cite{plimpton1995fast}.  \todo{skriv litt mer om hvorfor vi velger g(r)}

@@

\subsection{Likelihood}
The likelihood $\mathcal{L}$ of parameters $\theta$ of a model $M(\theta)$ given outcome $x$ is defined as $\mathcal{L} (M(\theta) | x) = P( x | M(\theta))$, where P is the probability distribution. In our model, we use the $\chi^2$ test to fit our one-dimensional measure from the data $\bf d$ with our model $\bf M(\theta)$. Pearson's $\chi^2$ is defined as 
\begin{equation}
  \chi^2 = \sum_i \frac{ \Big(d_i - M(\theta)_i \Big)^2}{M(\theta)_i^2},
\end{equation}    
and the likelihood is thus
\begin{equation}
\mathcal L \propto e^{-\chi^2}.
\end{equation}
For a model with $N$ free parameters, we build the full $N$-dimensional likelihood by letting random walkers map out the parameter space through the Monte Carlo Markov chain method (MCMC). For each step $M(\theta)_n$ in parameter space, we calculate the likelihood for this parameter configuration. Then, we create a new set of parameters for the next step $M(\theta)_{n+1}$ by choosing new parameters randomly from a $N$-dimensional sphere around the previous parameters. The model also chooses a new random seed. We calculate the likelihood at the new step and perform the MCMC test: Accept the new step if $\frac{\mathcal L_n}{\mathcal L_n+1} \ge 1$, or reject if the value is lower than a uniform random value $\in [0,1]$.

\section{Model verification and data analysis}
\begin{figure}[htb!]
\includegraphics[width=.45\textwidth]{model_test.png}
\caption{Simulated mock data used for the initial model testing analysis. Input parameters are $\nu=0.03$,$\tau=-0.02$ and $\psi=0.6$.}
\label{fig:mockdata}
\end{figure}
In order to verify our likelihood code, we perform a model verification test on a mock data set with known input values. In this example, we only use standard simplex noise with three free parameters: the persistence $\psi$, threshold $\tau$ and scale $\nu$. We simulate a data set of $159 ^3 Å$ of $1 000 000$ atoms, and perform the cut-off given input parameters $\Omega=3$, $\Psi = 0.6$, $\tau=-0.2$, $\nu=0.03$. The resulting 3D structure is depicted in figure \ref{fig:mockdata}.

@@

@@

\section{Results}
%\begin{figure}
%\includegraphics[width=.49\textwidth]{sio_porous.png}
%\caption{Simulated silica created with LAMMPS. The box is $159^3 \AA$, with $100 000$ particles, having been created by cooling an expanding system until quenching was achieved. Note the similarities between the pores here and in figure \ref{fig:mockdata}.}
%\label{fig:si2data}
%\end{figure}
\begin{figure*}
\includegraphics[width=.99\textwidth]{results_porous_full.png}
\caption{Results from the analysis of a simulated $159^3$ \AA $SiO^2$ nanoporous data set. The upper panel displays the marginalised 1D posteriors for each parameter, while the lower panel depicts the marginalised 2D posteriors. Best fit parameters from the maximum 3D likelihood suggest $\tau=-0.2$, $\psi=0.8$ and $\nu=0.3$.}
\label{fig:porous_results1}
\end{figure*}

We now perform a parameter fitting analysis on realistic simulated nanoporous media. Using LAMMPS \cite{plimpton1995fast}, we simulate a $100^3$ \AA box of 100K $SiO^2$ silica particles at an initial temperature of $4500 K$, using the Vashishta-potential. After an initial period, we expand the box to $159^3 \AA$ and start cooling until quenching is achieved at about $300K$. The resulting nanoporous media is depicted in figure \ref{fig:sio2data}, with a porosity of $75\%$.

With the model described in \ref{eq:noisemodel1}, we perform a full likelihood analysis of the data set with the three free parameters $\tau$, $\nu$ and $\psi$ using $g(r)$ as the measure. The resulting marginalised posteriors are depicted in figure \ref{fig:porous_results1}, with best-fit parameters $\tau=-0.2$, $\psi=0.8$ and $\nu=0.3$ is extracted from the full 3D likelihood. We also display the radial pore size distribution $g(r)$ for the input data set, a best-fit parameter data set and a bad-fit data set in figure \ref{fig:gofr1}, to illustrated the sensitivity of $g(r)$ given the model parameters. Note that we are not able to accurately reproduce the structures of the $SiO^2$ $g(r)$ with our current model, but hope that we during future work will be able to produce another model that provides a better fit with this measure.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract} 
In this paper, we aim to develop a method for producing statistically correct simulations of nanoporous media without the use of conventional Molecular Dynamics Simulator (MDS). In short, the proposed model is a binary function that returns either true if the input position is in a void / pore, or false if within a wall. By using procedural noise, we are not restricted to generating media within a limited box, as the noise functions is continuous on all scales. It is therefore possible to validated pores at theoretically arbitrary range. In addition, this method is extremely fast, requiring very little computational power to calculate whether a particle is within a pore or not. After presenting such a model, we choose a method to measure the structural properties of nanoporous media that are created either through regular methods (e.g. LAMMPS) or with our method. The measure we selected is the radial distribution function (RDF) $g(r)$, which is regarded as the de-facto standard method for capturing the structural properties of porous media. Using this measure, we define a full likelihood framework for estimating model parameters given any data set, and continue by validating our own framework by analysing procedural noise data sets with known input parameters. Finally, we perform a full likelihood analysis of silica SiO2 data set simulated in LAMMPS, and show that our model correctly reproduces several of the internal properties of the data set, including $g(r)$, porosity and surface area.  \todo{Abstract boer tenkes noe gjennom: faa frem hovedpoenger.}

@@

%During recent years, increase in computational power has allowed for a vast improvement  
\todo{Skriv en fin intro! Ogsaa om nanoporous media}
Properties of structural patterns that regularly occur in nature are often investigated either through direct observations or through simulations of theoretical models. Model simulations typically require a large computational framework to accurately reproduce the observed effects, while experimental data can be both expensive and hard to obtain, in addition to being riddled with artefacts and instrumental errors.

Procedural generation of physical structures have several advantages over conventional simulations. For example, when producing simulations of nanoporous silica, we are required to calculate and propagate the physical properties of each particle for every time-step throughout the simulation, using a computational framework such as LAMMPS \cite{plimpton1995fast}. This is both time-consuming, and the computational cost can sometimes be be quite large. In addition, the simulation is constrained to a box, so particles are not allowed to move outside the defined boundaries. On the other hand, using procedural noise to carve out nanoporous media is both extremely fast and computationally cost-efficient, in addition to being continuous on all scales. This means that it is possible to simulate

%In addition, by creating a model that yields properties that are statistically consistent with observations, we are ar

In this paper, we aim to investigate whether procedural methods are suitable for producing statistically correct 3D physical structures, focusing on nanoporous media. In a sense, we are combining well-known methods from computer games and real-time graphics-oriented software with experimental observations and simulations. Procedural methods for automated generation of computer game content have been around since the early 1980's, and in those early days algorithms were less advanced, often yielding plain, simple geometrical structures. Nowadays, almost every single computer game produced will contain some level of automated generated content, from textures to level design to landscapes and gameplay.

We do not claim that procedural noise methods completely reproduces all structural properties of nanoporous silica, but show that our approach can be used to efficiently create materials that are structurally similar to simulations. Naturally, this all depends on the statistics used for comparing the two models, which in this case is based on the radial distribution function (RDF) $g(r)$.

%\section{Method}
\section{Procedural content generation}
By procedural content generation, we generally mean a mathematical algorithm that uses limited input data to create complex, sometimes infinite repeating patterns. A firm example would be fractals, such as the Julaia/Mandelbrot [ref mandelbrot] set, where one tests whether the complex quadratic recurrence equation $z_{n+1} = z_n^2 +C$ for $z \in \mathbb C$ converges for each position in a 2D complex plane. With this seemingly simple equation, the level of complexity generated on the boundary between convergence and divergence is infinite, with self-similar patterns repeating themselves to arbitrary detail.
\begin{figure}
\includegraphics[width=.5\textwidth]{orphancity.png}
\caption{Example of city building using procedural methods: Landscape, city infrastructure, buildings and all textures are all automatically generated using procedural methods. Created by one of the authors. \todo{Denne er jo med bare for moro skyld, boer jo fjernes} }
\label{fig:orphancity}
\end{figure}

Another example, widely used in contemporary game content generation, is having prefabs (pre-fabricated 3D models such as doors, walls, tiles and roof segments) automatically being stitched together to create buildings. When designing a city, instead of creating each building manually, you only need specify the position, area and height of a structure. Then, by using an algorithm that combines these prefabs to render the building, it is possible to create cities of arbitrary size with a single command. Naturally, it is also possible to create procedural methods that design the layout and placement of these buildings. See figure \ref{fig:orphancity} for an example of such a city generator tool, developed by one of the authors of this paper.

A much simpler example would be generating a 1 (or 2)-dimensional landscape by choosing random points on a grid and performing a spline interpolation between them to create the illusion of a smooth landscape. In fact, this is the very essence of procedural noise, which we are focusing on in this paper, and the first person to create such a fast algorithm for N-dimensional spline interpolation between pseudo-random points was Ken Perlin.

\subsection{Perlin/Simplex noise}
\label{sec:perlin}

Procedural algorithmic methods for creating realistic-looking
fractal-like patterns have been around for more than 30 years. The most famous of these
algorithms is Perlin noise, first developed by Ken Perlin in 1982 for
use in the movie ``Tron''.

Perlin noise is a type of pseudo-random gradient noise that mimics a Gaussian field, but at low computational cost. An improved version of Perlin noise is called simplex noise, which has several advantages over standard Perlin noise in terms of scalability and speed. Perlin noise was developed in response to the lack of detail 1980's contemporary graphics, where polygonal surfaces often were bland and left without any detail. In response to this problem, Perlin noise was developed in order to continuously perturb the light bouncing off flat surfaces, effectively yielding bump mapping and producing a much more realistic image.

An overview of the current state of procedural noise functions can be
found in \cite{lagae:2010}.  Perlin noise \citep{perlin:2002} 
is a band limited repeatable pseudo-random function
from $\mathbb R^n \to \mathbb R$ that is computationally fast, and has
properties which make it suitable for creating patterns that mimic
nature. Perlin noise is an approximation to Gaussian filtered noise
implemented as a pseudo-random spline. The algorithm uses a predefined
grid of gradients pointing in a random direction, where for any point
within a grid, we interpolate the gradients from the four corners. The
algorithm is as follows:
\begin{itemize}
  \item[1.] For a point, obtain the four closest gradients in the grid.
   \item [2.] For each of the gradients, calculate the dot product
     between the gradient and a vector defined by the  the distance
     between the the point and the current grid corner. 
   \item [3.] Instead of linear interpolation $F(t) = t$, use a fade function to produce smoother interpolated values. Perlin noise typically uses the following fade function: $F(t) = t ^3 (t  (6 - 15) + 10)$.
    \item [4.] Obtain the final value by linearly interpolating between the x-axis and
      use this result to linearly interpolate the y-axis. 
\end{itemize} 
Simplex noise works in a similar manner, but requires fewer
computational steps, scales better in higher ($\ge 3$) dimensions and has fewer 
anisotropic properties.

\subsection{Properties of Perlin/Simplex noise}
\begin{figure}
\includegraphics[width=.5\textwidth]{noise.eps}
\caption{Perlin noise (left) versus simplex noise (right). Note the
  straight x-y pattern in the single upper Perlin mode and the less
  anisotropic simplex mode. Lower part: combining octaves tend
  to remove these patterns.}
\label{fig:noise}
\end{figure}

\begin{figure}
\includegraphics[width=.5\textwidth]{noise_comparison.png}
%\mbox{\epsfig{file=perlinstats.eps, width=\linewidth}}
\caption{Normalized distributions of Perlin noise and simplex noise compared with a
  Gaussian distribution. Note how both noise functions produce
  near-Gaussian, but slightly different distributions.}
\label{fig:properties}
\end{figure}
Standard Perlin noise is known for having several anisotropic features
\citep{lagae:2010}. Perlin noise also gives rise to a ``textile''
pattern in the $x$ and $y$-directions, which is evident from the upper
left part of figure \ref{fig:noise}, depicting 2D Perlin noise in
a single octave. The upper-right part shows the same node for simplex noise,
with less obvious patterns. However, when combining several octaves and adding a random shift to the coordinates,
as depicted in the two lower parts, these anisotropic patterns tend to cancel
out.  In addition, we have calculated the statistical distributions for Perlin and simplex noise
with $\sigma = 0.3$, and have plotted the results in figure \ref{fig:properties}.
Each procedural noise method has distributions that are close to normal, 
but with some asymmetrical properties.

\subsection{Multiple octaves: creating patterns}
\label{sec:octaves}
The different octaves from procedural noise functions can be combined
to produce a wide range of nature-like patterns. A generic
N-dimensional procedural function $\Phi(\vec x): \mathbb R^N \to
\mathbb R$ could be expressed as
\begin{equation}
\label{eq:procedural}
 \Phi(\vec x) = \Theta \Big(\sum_k^\omega \kappa (k) P(  f_s (\vec x,k) )) \Big),
 \end{equation}
where $f_s(\vec x,k)$ describes the frequency scale modifier (e.g. $f_s(\vec x,k) =
k\vec x$), $\omega$ is the amount of octaves (modes), $\kappa(k)$ is the octave amplitude (e.g. $\kappa(k) = 1/k$)
and $\Theta(x)$ an overall modifier (e.g. $\Theta(x) = x$ or
$\Theta(x) = \frac{1}{x}$). In a sense, $\kappa(k)$ is similar to Fourier coefficients. 
In its simplest case, summing directly over the octaves with amplitude $1/k$ results in 
\begin{equation}
\label{eq:perlinlinear}
 \Phi(\vec x) = \sum_k^\omega \frac{1}{k} P( 2k\vec x).
\end{equation}
In 1 dimension, equation \ref{eq:perlinlinear} with 1, 3 and 30 octaves is depicted in figure \ref{fig:1dperlin}, while the 2D procedural noise pattern depicted in the lower
part of figure \ref{fig:noise}. Finally, since simplex noise both scales better with higher dimensions and produces less asymmetric effects, we choose to use this noise method over regular Perlin noise in our method.

\begin{figure}
\includegraphics[width=.5\textwidth]{1d_perlin.png}
\caption{1D simplex noise with 1 (blue), 3 (green) and 30 (red) octaves. }
\label{fig:1dperlin}
\end{figure}

\section{Model}
In this paper, we are interested in developing a model using simplex noise to reproduce nanoporous structures that are as statistically close to actual simulations of porous media. For a given data set and a model, we wish to estimate the parameters that best represent the data. This is performed through a full likelihood analysis, where we in addition validate the model by comparing the radial distribution function $g(r)$, porosity and surface area of the system. The simulated system is confined to a cube of size $159 \AA^3$, but since the procedural noise function is continuous on all scales, the box could be of arbitrary size.

\begin{figure*}
\includegraphics[width=.95\textwidth]{model_examples.png}
\caption{Examples of 3D simplex noise structures in nanoporous media. Left image: low persistence (large scale dominates), middle image: medium persistence (small scales starts dominating) and right image: high threshold (more atoms are discarded). }
\label{fig:model_example}
\end{figure*}
We construct a generic sum over procedural noise octaves with the expression: 
\begin{equation}
  M_1(\bar \theta) = M_1(\vec x, \psi, \Omega, \nu) = \sum_{\omega=0}^{\Omega} \psi^{\omega -1}   N\big((\vec x + \sigma \cdot \omega)\nu \cdot 2^{\omega-1} \big),
\label{eq:noisemodel1}
\end{equation}
where $\Omega$ is the total amount of octaves, $\psi$ the persistence (scale falloff), $N$ a generic $N$-dimensional noise function, $\sigma$ a predefined random vector that removes anisotropic artefacts and $\nu$ the frequency modifier representing the initial largest scales in the model. For simplicity, $\bar \theta$ here denotes the set of model parameters. For each mode, the frequency is thus multiplied, while the overall amplitude is damped by a given persistence. In the end, we simulate a large set of bulk particles without any walls, and use the noise function to remove particles that are not in a procedural void given by a threshold $\tau$. Three examples of this model are depicted in figure \ref{fig:model_example}, where we have used $\Omega=6$ octaves and the scale is $\nu=0.01$ and cut-off $\tau=0$. The left image depicts low persistence ($\psi = 0.0$), where only the first octave dominates (corresponding to large scales). The middle picture shows the same parameters with a higher persistence $(\psi = 0.5)$, where smaller scales become more dominating. In the rightmost picture, the effect of the cut-off threshold $\tau = -0.3$ is shown.

@@

Ideally, the model measure should pick up all geometrical properties that are relevant for the physical study of interest. For instance, pore size distribution and porosity are import for studying transport in porous media. Another such measure is the Distance-To-Atom (DTA). The method used for calculating porosity of a media is defined as in \cite{gelb1998characterization}, and we obtain the surface area of a porous media by using marching cubes triangulation to sum the surface area of the triangles.

In this paper, we choose to use the radial distribution function $g(r)$ as defined and implemented in LAMMPS \cite{plimpton1995fast}. The $g(r)$ (or pair correlation function) describes how density varies as a function of distance from a reference particle. Hereafter, $g_d(r)$ denotes the RDF calculated from simulated or experimental data, while $g_{M(\bar \theta)}(r)$ represents the RDF based on the procedural noise model $M(\bar \theta)$, where $\bar \theta$ are the set of model parameters.

\todo{skriv litt mer om hvorfor vi velger g(r)}

@@

\subsection{Data analysis and the Likelihood}

The posterior probability is defined as the probability of the parameters $\bar \theta$ given the evidence $d$, or $P(\bar \theta | d)$. Bayes' theorem states that 
\begin{equation}
P(\bar \theta | d) = \frac{P(d | \bar \theta)P(\bar \theta)}{P(d)} \propto P(d | \bar \theta)P(\bar \theta),
\end{equation}
where $P(d | \bar \theta)$ is called the Likelihood $\mathcal L$ (the probability of evidence given parameters) and $P(\bar \theta)$ is the prior (the beliefs about the quantity before some evidence is taken into account). $P(d)$ is called the evidence, and is the probability of the observations (here assumed to be 1). Since the prior is a constant, we can therefore assume that the posterior distribution is proportional to the likelihood, or $P(d | \bar \theta) \propto \mathcal L$. In other words, to calculate the posterior and obtaining the best fit model parameters, we need only to estimate the likelihood function, which is connected to the $\chi^2$ by 
\begin{equation}
\mathcal L \propto e^{-\chi^2}.
\end{equation}

In order to test whether two 3D data sets are statistically equivalent up to the model measure $g(r)$, we calculate the $\chi^2$ between the two one-dimensional RDFs. Pearson's $\chi^2$ is defined as 
\begin{equation}
  \chi^2 = \sum_r \frac{ \Big(g_d(r) - g_{M(\bar \theta)}(r) \Big)^2}{g_{M(\bar \theta)}(r)^2},
\end{equation}
so for low values of $\chi^2$, the better the fit between data and model (up to $g(r)$). For a single-parameter model, we could simply just iterate over the lone parameter to obtain the minimum $\chi^2$ (where the best-fit parameter value is), however, with increased number of parameters comes increased computational cost. If we divide the parameter space into $N$ bins, the number of $\chi^2$ computations scales as $N^d$, where $d$ is the number of model parameters. To be able to build the full $\chi^2$ (and likelihood) parameter landscape, we there need to use Monte Carlo Markov chain methods (MCMC).

We are now interested in maximising the likelihood $\mathcal L$ as opposed to minimising the $\chi^2$, with the goal of mapping out the full $d$-dimensional likelihood space for the model parameters. We build the $d$-dimensional likelihood by letting random walkers traverse the parameter space guided by the MCMC test: for each step $M_n(\bar \theta)$ in parameter space, we calculate the likelihood for this parameter configuration. Then, we create a new set of parameters for the next step $M_{n+1}(\bar \theta)$ by choosing new parameters randomly from a $d$-dimensional sphere around the previous parameter composition. The model also selects a new random seed. We calculate the likelihood at the new step and perform the MCMC test: accept the new step if $\frac{\mathcal L_n}{\mathcal L_n+1} \ge 1$, or reject if the value is lower than a uniform random value $\in [0,1]$.

Finally, we obtain the posterior $P(\bar \theta | d) \propto \mathcal L$ by building a $d$-dimensional histogram of the likelihood using the stored positions of the parameters. For a sufficiently large amount of steps, the MCMC algorithm ensures that the posterior converges to the likelihood, as long at the surface does not contain unreachable areas such as discontinuous patches or an infinite amount of maxima (as is the case for the function $f(x)=sin(\frac{1}{x}$). Luckily, as we shall see, the likelihood surface for our models tend to turn out with a single, well-defined maximum.

\section{Results}
We now turn our attention to applying the MCMC framework on two kinds of data. First, we wish to test and verify the code by performing a full data analysis on mock data with known input parameters. This enables us to investigate the posterior distribution, and determine whether any potential problems could arise with the likelihood landscape, such as radom walkers trapped in local minima. We then continue with a full analysis of simulated SiO2 data, and see whether the resulting best-fit parameter data set is consistent with the SiO2 data set.

\subsection{Model verification: Parameter estimation of mock data}
\begin{figure}[htb!]
\includegraphics[width=.45\textwidth]{model_test.png}
\caption{Simulated mock data used for the initial model testing analysis. Input parameters are $\nu=0.03$,$\tau=-0.02$ and $\psi=0.6$.}
\label{fig:mockdata}
\end{figure}
In order to verify our likelihood code, we perform a model verification test on a mock data set with known input values. In this example, we only use standard simplex noise with three free parameters: the persistence $\psi$, threshold $\tau$ and scale $\nu$. We simulate a data set of $159 ^3 Å$ of $\sim 255 000$ atoms, and perform the cut-off given input parameters $\Omega=3$, $\Psi = 0.6$, $\tau=-0.2$, $\nu=0.03$. The resulting 3D structure is depicted in figure \ref{fig:mockdata}.

@@

Note that there seems to be no problems with the posterior distribution in terms of local minima or multiple maxima.

@@

\subsection{Parameter estimation of simulated SiO2 silica}
%\begin{figure}
%\includegraphics[width=.49\textwidth]{sio_porous.png}
%\caption{Simulated silica created with LAMMPS. The box is $159^3 \AA$, with $100 000$ particles, having been created by cooling an expanding system until quenching was achieved. Note the similarities between the pores here and in figure \ref{fig:mockdata}.}
%\label{fig:si2data}
%\end{figure}
\begin{figure*}
\includegraphics[width=.99\textwidth]{results_porous_full.png}
\caption{Results from the analysis of a simulated $159^3$ \AA SiO2 nanoporous data set. The upper panel displays the marginalised 1D posteriors for each parameter, while the lower panel depicts the marginalised 2D posteriors. Best fit parameters from the maximum 3D likelihood suggest $\tau=-0.2$, $\psi=0.8$ and $\nu=0.3$.}
\label{fig:porous_results1}
\end{figure*}

We now perform a parameter fitting analysis on realistic simulated nanoporous media. Using LAMMPS \cite{plimpton1995fast}, we simulate a $100^3\AA$  box of 100K SiO2 silica particles at an initial temperature of $4500 K$, using the Vashishta-potential. After an initial period, we expand the box to $159^3 \AA$ and start cooling until quenching is achieved at about $300K$. The resulting nanoporous media is depicted on the right side in figure \ref{fig:porous_vs_model}, with a porosity of $75\%$. \todo{Skrive mer om lagingen?}

With the model described in equation \ref{eq:noisemodel1}, we perform a full likelihood analysis of the data set with the three free parameters $\tau$, $\nu$ and $\psi$ using $g(r)$ as the measure. The resulting marginalised posteriors are depicted in figure \ref{fig:porous_results1}, with best-fit parameters $\tau=-0.2$, $\psi=0.8$ and $\nu=0.3$ is extracted from the full 3D likelihood. We also display the RFD $g(r)$ for the input data set, a best-fit parameter data set and a bad-fit data set in figure \ref{fig:gofr1}, to illustrated the sensitivity of $g(r)$ given the model parameters. Note that we are not able to accurately reproduce the structures of the SiO2 $g(r)$ with our current model, but hope that we during future work will be able to produce another model that provides a better fit with this measure.
